A Tale of Two Transects

### 2019-12-05 writing/transect_paper.Rmd
Don't refine MAFFT alignment, this takes FOREVER (and is throttled to 8 cores)

base.dir <- regmatches(getwd(), regexpr(".*oueme-fungi-transect", getwd()))
if (opts_chunk$get("dev") == "tikz") {
  # styles$Expression <- styles$LaTeX.Expression
  # styles$Name <- styles$LaTeX.Name
  # styles$Short.Name <- styles$LaTeX.Short.Name
  # styles$Unit <- paste0("\\si{", styles$LaTeX.Unit, "}")
}
write_bib(c("base", "vegan", "permute"), file = "R.bib", tweak = TRUE)
```
# Introduction
Send this outline by Thursday Dec 5
Meet Monday Dec 2 about

@@

## Turnover scale in ECM forests
New ecosystem type

@@

Amplicons were purified using SPRI beads [Vesterinen et al 2016] and quantified fluorometrically using PicoGreen dsDNA (Quant-iT) fluorescent indicator dye on a Infinite F200 plate spectrofluorometer (Tecan) according to the manufacturer's protocol.
100 ng of DNA from each sample (or the total PCR product if less than 100 ng) was pooled into two libraries each for long and short amplicons.
Each library was sequenced using a Pacific Biosciences RS II sequences.
Short amplicon libraries were sequenced on two SMRT cells each, while long amplicon libraries were sequenced on four SMRT cells each.

@@

Circular consensus sequence (CCS) basecalls were made using the `ccs` command from SMRTtools version 5.0.1  [@pacificbiosciences2019].
The resulting sequences were demultiplexed and sequencing primers were removed using `cutadapt` version 1.18 [@martin2011].

@@

Each of the extracted regions was independently filtered for length (table **X**) and a maximum of three expected errors.
Sequences were then dereplicated and denoised into amplicon sequencing variants (ASVs) using DADA2 version `r packageVersion("dada2")`.
The error model for DADA2 denoising was fit using only the 5.8S RNA region.
Independent error models were fit for each sequencing run (i.e., long *vs.* short amplicons, **different sequencing technologies**).
Chimeras within each region were removed using `removeBimeraDenovoTable` from DADA2.

@@

New taxonomic annotations of the Ribosomal Data Project's LSU fungal training set (RDP) version 11.5 [@cole2014] and Warcup ITS training set [@wang2007] were created, conforming to the taxonomic classification system used in the Unite database version 8 [@nilsson2019a].
In particular, the classification for fungi was according to @tedersoo2018, and for non-fungal eukaryotes was according to the proposed system of @tedersoo2017c.
Although the latter system is not formally published, it seems to be in use for non-fungal eukaryotes in the Unite database, and it is a system with only purportedly monophyletic taxa and a uniform set of taxon ranks.
These make it more appropriate for sequence-based taxonomic assignment algorithms than more accepted classification systems such as @adl2019.
FASTA format files of the re-annotated RDP and Warcup training sets are available at (*somewhere*).

Taxonomic assignment was performed to genus level on the ITS and LSU regions using Unite/Warcup and RDP, respectively, as taxonomic references.
For each region/reference combination, taxonomy was assigned using three algorithms:
the RDP Naïve Bayesian Classifier (NBC) as implemented in DADA2;
SINTAX [@edgar2016a] as implemented in VSEARCH v2.9.1 [@rognes2016];
and IDTAXA [@murali2018].
Each ASV thus was thus given nine taxonomic assignments (three region/reference combinations $\times$ three algorithms).

@@

Long amplicon ASVs were aligned as RNA using DECIPHER [@wright2015].
As attempt was made to refine this alignment using Pasta [@mirarab2014a], but the resulting maximum liklihood (ML) trees did not improve on the ML achieved using the original alignment from DECIPHER.
Since ML is the optimization criterion used by Pasta, the original DECIPHER alignment was retained.
A final ML tree was produced using RAxML version 8.2.12 [@stamatakis2014] using the GTR+GAMMA model and rapid bootstrapping with the MRE_IGN stopping criterion.

>>>>

base.dir <- regmatches(getwd(), regexpr(".*oueme-fungi-transect", getwd()))
if (opts_chunk$get("dev") == "tikz") {
  # styles$Expression <- styles$LaTeX.Expression
  # styles$Name <- styles$LaTeX.Name
  # styles$Short.Name <- styles$LaTeX.Short.Name
  # styles$Unit <- paste0("\\si{", styles$LaTeX.Unit, "}")
}
write_bib(c("base", "vegan", "permute"), file = "R.bib", tweak = TRUE)
```
# Introduction

@@

## Turnover scale in ECM forests

- Meta analysis @Bahram2013
    -

@@

Amplicons were purified using SPRI beads [Vesterinen et al 2016] and quantified fluorometrically using PicoGreen dsDNA (Quant-iT) fluorescent indicator dye on a Infinite F200 plate spectrofluorometer (Tecan) according to the manufacturer's protocol.
100 ng of DNA from each sample (or the total PCR product if less than 100 ng) was pooled into two libraries each for long and short amplicons.
Each library was sequenced using Single Molecule Real Time (SMRT) sequencing on a Pacific Biosciences RS II sequencer at the Uppsala Genome Center (UGS) at SciLifeLab in Uppsala, Sweden.
Short amplicon libraries were sequenced on two SMRT cells each, while long amplicon libraries were sequenced on four SMRT cells each.

Additionally the short amplicon libraries were combined and sequenced using an Ion S5  (Ion Torrent) sequencer using one 520 chip at UGS.

@@

Circular consensus sequence (CCS) basecalls for SMRT sequences were made using `ccs` version 3.4 [@pacificbiosciences2019] using the default settings.
The resulting sequences were demultiplexed and sequencing primers were removed using `cutadapt` version 1.18 [@martin2011].
Sequencing primers were similarly removed from the Ion Torrent sequences, but interference between the tagged gITS7 primers and the Ion XPress tags used in library prep made full demultiplexing of the Ion Torrent sequences impossible.

@@

Each of the extracted regions was independently filtered for length (table **X**) and a maximum of three expected errors.
Sequences were then dereplicated and denoised into amplicon sequencing variants (ASVs) using DADA2 version `r packageVersion("dada2")`.
The error model for DADA2 denoising was fit using the 5.8S RNA region for long amplicons, and using the entire read for short amplicons.
Independent error models were fit for each sequencing run (i.e., long *vs.* short amplicons, different sequencing technologies).
Chimeras within each region were removed using `removeBimeraDenovoTable` from DADA2.

@@

Taxonomic annotations of the Ribosomal Data Project's LSU fungal training set (RDP) version 11.5 [@cole2014] and Warcup ITS training set [@wang2007] were mapped to the taxonomic classification system used in the Unite database version 8 [@nilsson2019a].
In particular, the classification for fungi was according to @tedersoo2018, and for non-fungal eukaryotes was according to the proposed system of @tedersoo2017c.
Although the latter system is not formally published, it seems to be in use for non-fungal eukaryotes in the Unite database.
Additionally, it is a system with only purportedly monophyletic taxa and a uniform set of taxon ranks, which make it more appropriate for sequence-based taxonomic assignment algorithms than more accepted classification systems such as @adl2019.
FASTA format files of the re-annotated RDP and Warcup training sets are available at (*somewhere*).

Taxonomic assignment was performed to genus level on the ITS and LSU regions using Unite/Warcup and RDP, respectively, as taxonomic references.
For each region/reference combination, taxonomy was assigned using three algorithms:
the RDP Naïve Bayesian Classifier (NBC) as implemented in DADA2;
SINTAX [@edgar2016a] as implemented in VSEARCH v2.9.1 [@rognes2016];
and IDTAXA [@murali2018].
Each ASV thus was thus given up to nine taxonomic assignments (three region/reference combinations $\times$ three algorithms).

@@

Unique LSU regions from long amplicon ASVs were aligned as RNA using DECIPHER [@wright2015] with up to 10 iterations of progressive alignment and conserved secondary structure calculation and 10 refinement iterations.
This alignment was truncated at a position after the D3 region corresponding to base 907 of the *Saccharomyces cerevisiae* S288C reference sequence for LSU, because several sequences had type 1 introns after this position [@holst-jensen1999].
Full length long amplicon ASVs (including ITS1, 5.8S, and ITS2 regions) were aligned and truncated in the same way.

An attempt was made to refine the long amplicon alignment using Pasta [@mirarab2014a], but the resulting maximum likelihood (ML) trees did not improve on the ML achieved using the original alignment from DECIPHER.
Since ML is the optimization criterion used by Pasta, the original DECIPHER alignment was retained.

ML trees were produced using RAxML version 8.2.12 [@stamatakis2014] using the GTR+GAMMA model and rapid bootstrapping with the MRE_IGN stopping criterion.
The LSU tree was not constrained topologically, but the long amplicon tree was constrained by the result for the LSU tree.

Short amplicon ASVs without a perfect match to one of the long amplicon ASVs were added to the long amplicon alignment using the `--add` option of MAFFT version 7.453 [@katoh2012] with a 10-mer based guide tree and two iterations.
The short ASVs were then placed on the long amplicon tree using EPA-NG version 0.3.6 [@barbera2019].
A "grafted" tree including the short amplicon ASVs in the most probable positions from the EPA-NG results was produced using GAPPA version 0.5.1 [@czech2019a].
When more than one short amplicon ASV was placed in the same position on the tree, GAPPA by default groups all of these sequences together in a polytomy.
The branch leading to each of these polytomies was deleted by setting its length to zero and then using the `d2multi` function from the `ape` package in R.
The resulting tree was used as a guide tree to construct a final ML tree in RAxML, using the GTR+GAMMA model, in order to resolve relationships between different sequences assigned to the same branch.
No bootstrapping was performed on the final tree.
